Effective Dynamic Models of Metabolic Networks

### 2016-03-09 Paper_v1.tex
Draft1
Initial JV edits

\section{Introduction}
Metabolic networks for simple organisms such as \textit{E. coli} contain throusands of reactions that are highly interconnected. 
These organisms have the potential to be engineered to produce valuable products.
Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products. 
Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}. 
Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA\cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM)\cite{2008_kim_varner_ramkrishna_BiotechProg}.

Constraint based models of metabolic pathways, such as EM, FBA and MFA, attempt to describe intracellular metabolism using the stoichiometry of biochemical networks and a pseudo-steady state assumption\cite{2010_orth_NatBiotech}.
EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoihiometrically and thermodynamically feasible. 
However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space. 
Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective. But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell. 
HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.\
DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
Despite these advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.   
Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.

@@

\section{Results}
In the general case model, we demonstrated that the performance of the HCM FBA was equivalent to that of HCM EM.
In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}A), we calculated 6 elementary modes and 3 FBA modes.  
Using the HCM EM modeling framework, concentration profiles were generated for the extracellular metabolites and artifical biomass (Fig. \ref {fig:model-fitting}B). 
The kinetic parameters of HCM FBA were varied to fit the concentration profiles of the HCM EM model. 
For this proof of concept model, HCM FBA was able to use fewer modes and kinetic parameters to mimic the simulation profiles of HCM EM.

In the anaerobic \textit{E. coli} case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model\cite{2008_kim_varner_ramkrishna_BiotechProg}.
The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
The simulated HCM EM concentration profiles of biomass, substrate and fermentation products (Fig. \ref {fig:ecoli}A (dashed line)) were generated using the kinetic parameters and elementary modes reported\cite{2008_kim_varner_ramkrishna_BiotechProg}. 
The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data (Fig. \ref {fig:ecoli}A).
HCM FBA had a slightly better fit on lactate than HCM EM.

@@

In the aerobic \textit{E. coli} case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
A metabolic network\cite{2007_schuetz_etal_MolSysBio} of 98 reactions and 60 metabolites with the biomass reaction from Palsson's \textit{E. coli} core model\cite{2006_Palsson_model} was used.
We calculated over 66,000 elementary modes for this network and determined it was futile to model. 
For the same network, FBA computed 29 modes. 
HCM FBA was simulated to fit cell density, glucose and acetate measurements for an aerobic \textit{E. coli} batch culture\cite{1994_varma_palsson_ApplEnvMicro} (Fig.\ref{fig:ecoli}B (solid line)).

A variance based sensitivity analysis on the aerobic \textit{E. coli} case identified essential modes that were influential on biomass yield. 
The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions (Fig. \ref{fig:sensitivity}).
We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.   
Five FBA modes were determined to be significant by an arbitrary threshold of 0.00218. 
The remaining 24 FBA modes were identified to be minor modes and insignificant. 
The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B (dotted line)) was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.

\begin{figure*}[!t]\centering
\includegraphics[width=0.95\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
\caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate. Sensitivity indicies computed for rate constants, saturation constants and enzyme parameters. The most critical parameters are associated with mode 1 representing aerobic growth. (Error bars represent 95\% confidence intervals).    
}
\label{fig:sensitivity}
\end{figure*}

@@

\noindent\subsubsection*{Description of the method}
The model was constructed using the hybrid cybernetic approach of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}. 
The abundance of extracellular species $i$ ($x_{i}$) is governed by:

\begin{equation}
	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)c \qquad{i=1,\hdots,\mathcal{M}} 
\end{equation}

where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of modes.
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and
$z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
The $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)$ term denotes the regulated uptake rate of mode $k$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable for enzyme activity ($v_{l}$), $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right) = \bar{r}_{l}v_{l}$, where $\mathbf{e}$ denotes the enzyme level for each mode which become additional state variables, $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector, and $\mathbf{s}$ denotes the substrate utilized. The reaction rate followed Michaelis-Menten kinetics. The abundance of the enzyme species is governed by:

\begin{equation}
	\frac{de_{l}}{dt}  = \alpha + r_{El}u_{l} - \left(\beta+r_{G}\right)e_{l}
\end{equation}

where $\alpha$ denotes the constant background enzyme synthesis, $r_{El}$ denotes the enzyme synthesis reaction and $u_{l}$ denotes the cybernetic control variable for enzyme synthesis. In this study, the enzyme synthesis reaction follows Michealis-Menten kinetics. The $\beta$ term denotes the degradation rate of enzyme, and $r_{G}$ denotes the growth rate through all modes. All initial conditions of the enzyme species were set to 0.9 for the anaerobic case and 0.8 for the aerobic case.

\begin{equation}
	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)
\end{equation}

where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$. 
The cybernetic control variables, $u_{l}$ and $v_{l}$, control the synthesis and activity of each enzyme for each mode using the matching and proportional laws\cite{2007_young_ramkrishna_BiotechProg}. 
They are formulated to maximize the overall substrate uptake rate through each mode.

\begin{equation}
	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} 
\end{equation}
\begin{equation}
	v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max_{l=1} z_{sl}\bar{r}_{l}}
\end{equation}

where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$. 
In the anaerobic case we followed the assumption of Young's thesis\cite{2005_Young} and Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg} that formate decomposition occurs only outside the network (applicable for strain GJT001). 
Therefore its enzyme activity was assumed to be at its maximum level of unity. 
The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}

@@

\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\ 
\alpha_i \leq w_i \leq \beta_i  \qquad
 \end{multlined}
\end{equation}

%\begin{alignat}{3}
%	\max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
 %	 \mathrm{Subject \; to:}
%	 \; \;  \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\
%	\alpha_i \leq w_i \leq \beta_i 
%\end{alignat}
where $\boldsymbol{c}$ is a vector of weights indicating how each flux contributes to the objective; $\alpha_i$ and $\beta_i$ are the lower and upper limits for the individual flux values, respectively. 
The stoichoimetric matrix $\mathbf{S}$ denotes the intracellular network where intracellular metabolism has a psuedo steady state approximation. 
$\mathbf{S_{x}}$ denotes the extracellular metabolic network and $\mathbf{b}$ denotes the exchange flux across the cell membrane. 
For each mode, the objective was set to maximize either the growth flux or each fermentation byproduct flux from a specified substrate.
Multiple modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either nonexistent or its maximum.
In the aerobic case, oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively. 
Each flux vector was normalized by the growth flux, otherwise by the specified objective flux.

@@

\end{document}

>>>>

\section{Introduction}
Metabolic networks for simple organisms such as \textit{E. coli} contain thousands of reactions that are highly interconnected.
These organisms have the potential to be engineered to produce valuable products.
Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.

Convex and constraints based analysis methods, such as EM, FBA and MFA,
describe intracellular metabolism using the stoichiometry of the biochemical networks and the pseudo-steady state assumption \cite{2010_orth_NatBiotech}.
EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoihiometrically and thermodynamically feasible.
However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective. But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.\
DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
Despite these advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.

@@

\section{Results}
The performance of HCM FBA was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}A), we calculated 6 elementary modes and 3 FBA modes.
Using the HCM EM modeling framework, concentration profiles were generated for the extracellular metabolites and artifical biomass (Fig. \ref {fig:model-fitting}B).
The kinetic parameters of HCM FBA were varied to fit the concentration profiles of the HCM EM model.
For this proof of concept model, HCM FBA was able to use fewer modes and kinetic parameters to mimic the simulation profiles of HCM EM.

The performance of HCM FBA was equivalent to HCM EM for a small model of anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
In the  case, we demonstrated that  has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
The simulated HCM EM concentration profiles of biomass, substrate and fermentation products (Fig. \ref {fig:ecoli}A (dashed line)) were generated using the kinetic parameters and elementary modes reported\cite{2008_kim_varner_ramkrishna_BiotechProg}.
The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data (Fig. \ref {fig:ecoli}A).
HCM FBA had a slightly better fit on lactate than HCM EM.

@@

HCM FBA outperformed HCM EM for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
In the  case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
A metabolic network\cite{2007_schuetz_etal_MolSysBio} of 98 reactions and 60 metabolites with the biomass reaction from Palsson's \textit{E. coli} core model\cite{2006_Palsson_model} was used.
We calculated over 66,000 elementary modes for this network and determined it was futile to model.
For the same network, FBA computed 29 modes.
HCM FBA was simulated to fit cell density, glucose and acetate measurements for an aerobic \textit{E. coli} batch culture\cite{1994_varma_palsson_ApplEnvMicro} (Fig.\ref{fig:ecoli}B (solid line)).

Variance based sensitivity of aerobic \textit{E. coli} metabolism identified modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions.
We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.
Five FBA modes were determined to be significant by an arbitrary threshold of 0.00218.
The remaining 24 FBA modes were identified to be minor modes and insignificant.
The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B (dotted line)) was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.

\begin{figure*}[!t]\centering
\includegraphics[width=0.85\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
\caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate. Sensitivity indicies computed for rate constants, saturation constants and enzyme parameters. The most critical parameters are associated with mode 1 representing aerobic growth. (Error bars represent 95\% confidence intervals).
}
\label{fig:sensitivity}
\end{figure*}

@@

\noindent\subsubsection*{HCM FBA method.}
The model was constructed using the hybrid cybernetic approach of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
The abundance of extracellular species $i$ ($x_{i}$) is governed by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)c \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of modes.
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and
$z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
The $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)$ term denotes the regulated uptake rate of mode $k$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable for enzyme activity ($v_{l}$), $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right) = \bar{r}_{l}v_{l}$, where $\mathbf{e}$ denotes the enzyme level for each mode which become additional state variables, $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector, and $\mathbf{s}$ denotes the substrate utilized. The reaction rate followed Michaelis-Menten kinetics. The abundance of the enzyme species is governed by:

\begin{equation}
	\frac{de_{l}}{dt}  = \alpha_{l} + r_{E,l}u_{l} - \left(\beta_{l}+r_{G}\right)e_{l}
\end{equation}
where $\alpha_{l}$ denotes the specific rate of constitutive enzyme synthesis for enzyme l, $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
The specific rate of enzyme synthesis was modeled using saturation kinetics.
All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.

\begin{equation}
	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)
\end{equation}
where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$.
The cybernetic control variables, $u_{l}$ and $v_{l}$, control the synthesis and activity of each enzyme for each mode using the matching and proportional laws\cite{2007_young_ramkrishna_BiotechProg}.
They are formulated to maximize the overall substrate uptake rate through each mode.

\begin{equation}
	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}}
\end{equation}
\begin{equation}
	v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max_{l=1} z_{sl}\bar{r}_{l}}
\end{equation}
where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
In the anaerobic case we followed the assumption of Young's thesis\cite{2005_Young} and Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg} that formate decomposition occurs only outside the network (applicable for strain GJT001).
Therefore its enzyme activity was assumed to be at its maximum level of unity.
The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}

@@

\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\
\alpha_i \leq w_i \leq \beta_i  \qquad
 \end{multlined}
\end{equation}
where $\boldsymbol{c}$ is a vector of weights indicating how each flux contributes to the objective; $\alpha_i$ and $\beta_i$ are the lower and upper limits for the individual flux values, respectively.
The stoichoimetric matrix $\mathbf{S}$ denotes the intracellular network where intracellular metabolism has a psuedo steady state approximation.
$\mathbf{S_{x}}$ denotes the extracellular metabolic network and $\mathbf{b}$ denotes the exchange flux across the cell membrane.
For each mode, the objective was set to maximize either the growth flux or each fermentation byproduct flux from a specified substrate.
Multiple modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either nonexistent or its maximum.
In the aerobic case, oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
Each flux vector was normalized by the growth flux, otherwise by the specified objective flux.

@@

\end{document}
\grid
